Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-29 manu-draft.Rmd
CL comments addressed

data(rswqdat)
data(rstrnpts)
data(rstrndat)
data(trnsum)

@@

Statistical assessments were evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners. Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below.

@@

A general summary of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline). After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April chlorophyll concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.  Conditions were relatively stable in August and September compared to the months prior. A quantitative description of these events follows.

@@

```{r}
nwq <- sum(wqsumtab[['N obs.']])
outrng1 <- wqsumtab %>% 
  filter(`Water quality variable` == 'Secchi (m)') %>% 
  pull(below) %>%  
  sum
outrng2 <- wqsumtab %>% 
  filter(`Water quality variable` != 'Secchi (m)') %>% 
  pull(above) %>%  
  sum
outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 1)

# percentages out of range
wqsumfl <- wqsumtab %>% 
  mutate(
    Area = ifelse(Area == '', NA, Area)
  ) %>% 
  fill(Area) 
chlabv <- wqsumfl %>% 
  filter(`Water quality variable` == 'Chl-a (ug/L)') %>% 
  select(Area, above) %>% 
  deframe %>% 
  round(0)
tnabv <- wqsumfl %>% 
  filter(`Water quality variable` == 'TN (mg/L)') %>% 
  select(Area, above) %>% 
  deframe %>% 
  round(0)
secblw <- wqsumfl %>% 
  filter(`Water quality variable` == 'Secchi (m)') %>% 
  select(Area, below) %>% 
  deframe %>% 
  round(0)
dosabv <- wqsumfl %>% 
  filter(`Water quality variable` == 'DO (% sat.)') %>% 
  select(Area, above) %>% 
  deframe %>% 
  round(0)
```

@@

From April through September 2021, `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).  Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others). The percentage of observations outside of the normal range varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1 located closest to teh discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in area 1, `r dosabv[['2']]`\% in area 2).  Physical parameters (salinity, temperature) were generally within range over the study period. Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the discharge measurements in Table \@ref(tab:stktab). Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).

@@

# all fo
foall <- rstrndat %>% 
  filter(date < maxdt) %>% 
  mutate(
    pa = bb > 0
  ) %>% 
  mutate(
    taxa = as.character(taxa), 
    taxa = case_when(
      grepl('^Halophila', taxa) ~ 'Halophila sp', 
      T ~ taxa
    )
  ) %>% 
  group_by(taxa) %>% 
  summarise(
    fo = sum(pa) / length(pa), 
    .groups = 'drop'
  ) %>% 
  mutate(
    fo = round(100 * fo, 1)
  ) %>% 
  deframe

# fo by month, area
fomo <- trnsum %>% 
  ungroup() %>% 
  mutate(
    foest = round(100 * foest, 1), 
    mo = month(date)
  )
fomoarea1 <- fomo %>% 
  filter(area == 'Area 1')
fomoarea3 <- fomo %>% 
  filter(area == 'Area 3')
fomoarea1ave <- fomoarea1 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 1)) %>% 
  deframe
fomoarea3ave <- fomoarea3 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 1)) %>% 
  deframe

@@

```{r}
fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>% 
  select(
    date = textBox6, 
    city = cOUNTYDataTextBox, 
    waterbody = lOCATIONDataTextBox,
    species = textBox18
  ) %>% 
  mutate(
    date = mdy(date),
    yr = year(date),
    week = floor_date(date, unit = 'week'), 
    week = factor(format(week, '%b %d')), 
    week = factor(week, levels = as.character(unique(week))), 
    city = gsub('\\s+$', '', city),
    city = gsub('^St\\s', 'St. ', city)
  ) %>% 
  filter(city %in% c('Tampa', 'St. Petersburg')) %>% 
  filter(date < maxdt)
fishwk <- fishdat %>% 
  filter(yr == 2021) %>% 
  group_by(week, city) %>% 
  summarise(cnt = n(), .groups = 'drop') %>% 
  group_by(city) %>% 
  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
  mutate(per = round(100 * cnt / sum(cnt), 1))
wkstpcnt <- fishwk %>% 
  filter(city == 'St. Petersburg') %>% 
  pull(cnt)
wktmpcnt <- fishwk %>% 
  filter(city == 'Tampa') %>% 
  pull(cnt)
```

The increase in *K. brevis* from April to July to bloom concentrations exceeding 10,000 cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in middle and lower Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021. Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.

@@

The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).  This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*).  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.  However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, total nitrogen and chlorophyll-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point were likely present (as noted by modelling efforts).  Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.  These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).

@@

@@

In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen effluent directly into lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters discharged from Piney Point. Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].  From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication and primary production in 2021 were comparatively different than recent prior years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.

@@

For seagrasses, both major bloom events produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.  These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red tide bloom in 2018.  Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

@@

Fisheries resources were undoubtedly impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, restrictions for catch and release only for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining are a serious threat to environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from gypstack release.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the effluent.  Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Although the characteristics of water in a gypstack can vary based on site specific criteria, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that can significantly affect coastal resources.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.

Limitations of our analyses are also important to note to inform future event-based monitoring and additional research.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.  The weight of evidence that 2021 is an anomaly relative to recent prior years provides justification that our interpretation is within the limits of what would be expected by more mechanistic descriptions.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling and fate of nitrogen from Piney Point.  There may likely never be a definitive link, but this is not necessary to support future actions that ensure that the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.  These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.

>>>>

data(rswqdat)
data(rstrnpts)
data(rstrndat)
data(trnsum)
data(tnanndat)

@@

Statistical assessments were evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners. For censored data (i.e., above or below detection), secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.

@@

A general summary of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline). After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April chlorophyll concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (filamentous cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in early to mid July, with bloom conditions not observed in the bay after July.  Conditions were relatively stable in August and September compared to the months prior. A quantitative description of these events follows.

@@

```{r}
nwq <- sum(wqsumtab[['N obs.']])
outrng1 <- wqsumtab %>% 
  filter(`Water quality variable` == 'Secchi (m)') %>% 
  pull(`% below`) %>%  
  sum
outrng2 <- wqsumtab %>% 
  filter(`Water quality variable` != 'Secchi (m)') %>% 
  pull(`% above`) %>%  
  sum
outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 0)

# percentages out of range
wqsumfl <- wqsumtab %>% 
  mutate(
    Area = ifelse(Area == '', NA, Area)
  ) %>% 
  fill(Area) 
chlabv <- wqsumfl %>% 
  filter(`Water quality variable` == 'Chl-a (ug/L)') %>% 
  select(Area, `% above`) %>% 
  deframe %>% 
  round(0)
tnabv <- wqsumfl %>% 
  filter(`Water quality variable` == 'TN (mg/L)') %>% 
  select(Area, `% above`) %>% 
  deframe %>% 
  round(0)
secblw <- wqsumfl %>% 
  filter(`Water quality variable` == 'Secchi (m)') %>% 
  select(Area, `% below`) %>% 
  deframe %>% 
  round(0)
dosabv <- wqsumfl %>% 
  filter(`Water quality variable` == 'DO (% sat.)') %>% 
  select(Area, `% above`) %>% 
  deframe %>% 
  round(0)
ammsum <- rswqdat
```

@@

From April through September 2021, `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).  Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others). The percentage of observations outside of the normal range varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1 located closest to teh discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in area 1, `r dosabv[['2']]`\% in area 2).  Physical parameters (salinity, temperature) were generally within range over the study period. Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1.  Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).

@@

# all fo
foall <- rstrndat %>% 
  filter(date < maxdt) %>% 
  mutate(
    pa = bb > 0
  ) %>% 
  mutate(
    taxa = as.character(taxa), 
    taxa = case_when(
      grepl('^Halophila', taxa) ~ 'Halophila sp', 
      T ~ taxa
    )
  ) %>% 
  group_by(taxa) %>% 
  summarise(
    fo = sum(pa) / length(pa), 
    .groups = 'drop'
  ) %>% 
  mutate(
    fo = round(100 * fo, 0)
  ) %>% 
  deframe

# fo by month, area
fomo <- trnsum %>% 
  ungroup() %>% 
  mutate(
    foest = round(100 * foest, 0), 
    mo = month(date)
  )
fomoarea1 <- fomo %>% 
  filter(area == 'Area 1')
fomoarea3 <- fomo %>% 
  filter(area == 'Area 3')
fomoarea1ave <- fomoarea1 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 0)) %>% 
  deframe
fomoarea3ave <- fomoarea3 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 0)) %>% 
  deframe

@@

```{r}
fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>% 
  select(
    date = textBox6, 
    city = cOUNTYDataTextBox, 
    waterbody = lOCATIONDataTextBox,
    species = textBox18
  ) %>% 
  mutate(
    date = mdy(date),
    yr = year(date),
    week = floor_date(date, unit = 'week'), 
    week = factor(format(week, '%b %d')), 
    week = factor(week, levels = as.character(unique(week))), 
    city = gsub('\\s+$', '', city),
    city = gsub('^St\\s', 'St. ', city)
  ) %>% 
  filter(city %in% c('Tampa', 'St. Petersburg')) %>% 
  filter(date < maxdt)
fishwk <- fishdat %>% 
  filter(yr == 2021) %>% 
  group_by(week, city) %>% 
  summarise(cnt = n(), .groups = 'drop') %>% 
  group_by(city) %>% 
  summarise(cnt = sum(cnt), .groups = 'drop')
wkstpcnt <- fishwk %>% 
  filter(city == 'St. Petersburg') %>% 
  pull(cnt)
wktmpcnt <- fishwk %>% 
  filter(city == 'Tampa') %>% 
  pull(cnt)
```

The increase in *K. brevis* from April to July to bloom concentrations exceeding 10,000 cells/L (based on FWC breakpoint) was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in middle and lower Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021. Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.

@@

The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).  This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms ((e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*, Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the total nitrogen in the initial discharge.  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.  However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, total nitrogen and chlorophyll-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point were likely present (as noted by modelling efforts).  Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.  These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).

@@

```{r}
# tnanndat from https://github.com/tbep-tech/load-estimates
tnld <- tnanndat %>% 
  filter(bay_segment == 'Lower Tampa Bay') %>% 
  filter(year > 2005 & year < 2021) %>% 
  group_by(year) %>% 
  summarise(
    tn_load = sum(tn_load)
  ) %>% 
  pull(tn_load) %>% 
  mean %>% 
  `/`(1.102) %>% 
  round(0)
```

@@

In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen effluent directly into lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters discharged from Piney Point. Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].  External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).  From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication and primary production in 2021 were comparatively different than recent prior years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.

@@

For seagrasses, both major bloom events produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018.  Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

@@

Fisheries resources were undoubtedly impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, restrictions for catch and release only for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining are a serious threat to environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from gypstack release.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the effluent.  Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Although the characteristics of water in a gypstack can vary based on site specific criteria, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that can significantly affect coastal resources.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.

Limitations of our analyses are also important to note to inform future event-based monitoring and additional research.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.  The weight of evidence that 2021 is an anomaly for both nitrogen loading and water quality response relative to recent prior years provides justification that our interpretation is within the limits of what would be expected by more mechanistic descriptions.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.  Long-term fate of nitrogen from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021. There may likely never be a definitive link, but this is not necessary to support future actions that ensure that the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.  These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
